---
title: "phenoscapR at Scale"
subtitle: "A 500,000-cell solid-organ cohort: vessels, immune cuffs, and parenchyma"
author: "R. Heller"
date: "`r format(Sys.Date())`"
output:
  pdf_document:
    toc: true
    toc_depth: 2
    number_sections: true
    fig_caption: true
    df_print: kable
fontsize: 10pt
geometry: margin=2.2cm
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>", echo = TRUE,
  message = FALSE, warning = FALSE,
  fig.width = 6.6, fig.height = 4.2, fig.align = "center", dpi = 150
)
options(width = 92)
library(phenoscapR); library(ggplot2)
source("data-raw/make_large_data.R")
has_deldir <- requireNamespace("deldir", quietly = TRUE)
has_uwot   <- requireNamespace("uwot", quietly = TRUE)
tt <- function(expr) round(system.time(expr)[["elapsed"]], 2)
timings <- list()
```

\newpage

# A half-million-cell synthetic cohort

This document scales the phenoscapR workflow to **500,000 cells across ten
tissue sections** to show that the kd-tree search engine keeps the spatial
statistics tractable on realistic cohorts.

The tissue is modelled as a **solid organ**:

* **CD31+ endothelial cells trace vessel paths** — these are the spatial
  *anchors*.
* **Immune cells** (T helper / cytotoxic / regulatory, B cells, macrophages)
  are **enriched perivascularly**, cuffed around the vessels, with a scattered
  minority.
* **Parenchyma** (PanCK+ epithelium) forms the bulk, punctuated by a few
  proliferative **Ki67-high tumour nests**.

```{r build}
timings$`Build object (500k cells)` <- tt(
  spe <- make_large_spatial(n_per_sample = 50000L, n_samples = 10L))
spe
```

```{r composition-tab}
spe@meta_data$phenotype <- spe@meta_data$phenotype_true
round(prop.table(table(spe$phenotype_true)), 3)
```

# Performance at scale

Every neighbour query below runs on the **full 500k cells** (sample-aware: the
engine never mixes sections). The single-window point-pattern statistics run on
one 50k-cell section.

```{r whole-cohort}
timings$`CellDensity (500k, r=30)` <- tt(spe <- CellDensity(spe, radius = 30))
timings$`FindNeighbours k=5 (500k)` <- tt(spe <- FindNeighbours(spe, k = 5L))
timings$`InteractionMatrix (500k, r=30)` <- tt(
  im <- InteractionMatrix(spe, radius = 30))
```

```{r one-section}
one <- spe[spe$sample_id == "section_01", ]
NCells(one)
timings$`RipleysK border (50k section)` <- tt(
  rk <- RipleysK(one, r_seq = seq(0, 200, length.out = 40), correction = "border"))
timings$`MoransI CD31 (50k section)` <- tt(
  mi <- MoransI(one, feature = "CD31", radius = 30))
timings$`NeighbourhoodEnrichment (50k, 99 perm)` <- tt(
  ne <- NeighbourhoodEnrichment(one, radius = 30, n_perm = 99L, seed = 1))
```

```{r timings-table, echo = FALSE}
data.frame(Step = names(timings),
           `Elapsed (s)` = unlist(timings),
           check.names = FALSE, row.names = NULL)
```

\newpage

# Tissue architecture

A single 50k-cell section. The CD31+ vessels and the immune cuffs around them
are visible directly in the phenotype map.

```{r cellmap, dev = "ragg_png", fig.height = 5.2, fig.cap = "One section (50,000 cells). Endothelial cells trace the vessels; immune cells cuff around them; epithelium fills the stroma."}
CellMap(one, pt_size = 0.35)
```

```{r cd31, dev = "ragg_png", fig.height = 4.8, fig.cap = "CD31 intensity isolates the vessel network — the spatial anchors of the tissue."}
FeaturePlot(one, features = "CD31", pt_size = 0.35)
```

# Vessels as spatial anchors

## Immune cells sit closer to vessels than parenchyma does

`CrossNNDistance()` measures, for each cell of a phenotype, the distance to the
nearest CD31+ endothelial cell. Perivascular immune populations have much
smaller distances than the uniformly distributed epithelium.

```{r crossnn, fig.height = 4.2, fig.cap = "Distance to the nearest vessel (CD31+ endothelium) by phenotype. Immune cells are perivascular; epithelium is not."}
phenos <- setdiff(unique(one$phenotype), "Endothelial")
dist_to_vessel <- lapply(phenos, function(p)
  data.frame(phenotype = p,
             d = CrossNNDistance(one, from = p, to = "Endothelial")))
dtv <- do.call(rbind, dist_to_vessel)

med <- tapply(dtv$d, dtv$phenotype, median)
dtv$phenotype <- factor(dtv$phenotype, levels = names(sort(med)))
ggplot(dtv, aes(phenotype, d, fill = phenotype)) +
  geom_boxplot(outlier.size = 0.2, linewidth = 0.3) +
  scale_fill_manual(values = PaletteDiscrete(nlevels(dtv$phenotype))) +
  labs(x = NULL, y = "distance to nearest vessel") +
  theme_minimal(base_size = 11) +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 30, hjust = 1))
```

```{r crossnn-med, echo = FALSE}
data.frame(phenotype = names(sort(med)),
           `median distance to vessel` = round(sort(med), 1),
           check.names = FALSE, row.names = NULL)
```

## The perivascular gradient

Binning every non-endothelial cell by its distance to the nearest vessel shows
the cuffing directly: close to CD31+ vessels the tissue is overwhelmingly
immune, and the immune fraction decays into the parenchyma with distance.

```{r gradient, fig.height = 3.8, fig.cap = "Fraction of cells that are immune as a function of distance to the nearest vessel. The perivascular zone is immune-dominated; the bulk parenchyma is not."}
imm <- c("T helper", "T cytotoxic", "T reg", "B cell", "Macrophage")
dd <- dtv[dtv$d <= 180, ]
dd$immune <- dd$phenotype %in% imm
br <- seq(0, 180, by = 15)
dd$bin <- cut(dd$d, br, include.lowest = TRUE)
grad <- data.frame(distance = head(br, -1) + 7.5,
                   immune_fraction = as.numeric(tapply(dd$immune, dd$bin, mean)))
ggplot(grad, aes(distance, immune_fraction)) +
  geom_area(fill = "#3366cc", alpha = 0.25) +
  geom_line(colour = "#3366cc", linewidth = 1) +
  geom_point(colour = "#3366cc", size = 1.6) +
  scale_y_continuous(labels = scales::percent, limits = c(0, NA)) +
  labs(x = "distance to nearest vessel (CD31+)",
       y = "cells that are immune") +
  theme_minimal(base_size = 11)
```

## Co-localisation with the endothelium

`NeighbourhoodEnrichment()` confirms it formally: immune phenotypes are
enriched as neighbours of endothelium (positive z), epithelium is not.

```{r enrich-endo, echo = FALSE}
endo <- ne[ne$to == "Endothelial" & ne$from != "Endothelial",
           c("from", "z_score")]
endo <- endo[order(-endo$z_score), ]
endo$z_score <- round(endo$z_score, 1)
endo
```

\newpage

# Phenotype interactions across the cohort

```{r interactions, fig.height = 5, fig.cap = "Interaction scores (log2 observed/expected) pooled per-sample across all 500k cells."}
InteractionPlot(im)
```

# Multi-scale clustering

```{r ripley, fig.height = 3.8, fig.cap = "Border-corrected Ripley's L on a 50k section; positive values indicate clustering driven by the vascular and tumour structure."}
with(rk, { plot(r, L, type = "l", lwd = 2, col = "#3366cc",
  xlab = "r", ylab = "L(r)", main = "Ripley's L (border-corrected, 50k cells)")
  abline(h = 0, lty = 2) })
```

```{r moran, echo = FALSE}
cat("Moran's I for CD31 (vessel autocorrelation):\n"); unlist(mi)
```

# Composition across sections

```{r composition, fig.height = 4, fig.cap = "Phenotype composition is consistent across the ten sections."}
CompositionPlot(spe, group_by = "sample_id") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

\newpage

# Marker space (dimensionality reduction)

PCA runs on all 500k cells; UMAP on a 30k-cell subsample for legibility.

```{r pca, dev = "ragg_png", fig.height = 4.4, fig.cap = "PCA of marker space (500k cells) coloured by phenotype."}
timings_pca <- tt(spe <- RunPCA(spe, n_pcs = 12))
set.seed(1)
sub <- spe[sample(NCells(spe), 30000L), ]
DimPlot(sub, reduction = "pca", colour_by = "phenotype", pt_size = 0.3)
```

```{r umap, dev = "ragg_png", eval = has_uwot, fig.height = 4.4, fig.cap = "UMAP of a 30k-cell subsample; immune, endothelial, and parenchymal compartments separate cleanly."}
sub <- RunUMAP(sub, dims = 12, n_neighbors = 20)
EmbeddingPlot(sub, reduction = "umap", colour_by = "phenotype", pt_size = 0.3)
```

# Session information

```{r session}
sessionInfo()
```
