---
title: "phenoscapR --- Full Capabilities Report"
subtitle: "From cell tables to niches, domains, and differential abundance"
author: "R. Heller and H. Witte"
date: "`r format(Sys.Date())`"
output:
  pdf_document:
    toc: true
    toc_depth: 2
    number_sections: true
    fig_caption: true
    df_print: kable
fontsize: 10pt
geometry: margin=2.2cm
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>", echo = TRUE,
  message = FALSE, warning = FALSE,
  fig.width = 6.6, fig.height = 4.2, fig.align = "center", dpi = 150
)
options(width = 92)
library(phenoscapR); library(ggplot2)
has_mclust <- requireNamespace("mclust", quietly = TRUE)
has_seurat <- requireNamespace("Seurat", quietly = TRUE) &&
  requireNamespace("SeuratObject", quietly = TRUE)
has_deldir <- requireNamespace("deldir", quietly = TRUE)
```

\newpage

# Overview

**phenoscapR** is a complete toolkit for single-cell spatial biology: it reads
multiplexed-imaging cell tables, runs quality control, normalisation, and
phenotyping, computes a rigorous suite of spatial statistics on a scalable
kd-tree engine, finds higher-order structure (cellular neighbourhoods and
spatial domains), tests differential abundance across conditions, and bridges
to the Seurat / SpatialExperiment ecosystem.

This report exercises the whole feature set on the bundled synthetic dataset
`phenoscapR_example` (a two-section tonsil assay), so every result is real and
reproducible.

```{r data}
data(phenoscapR_example)
obj <- phenoscapR_example
obj@meta_data$phenotype <- obj@meta_data$phenotype_true
obj <- NormaliseData(obj, "zscore")
obj
```

# Core workflow

```{r pipeline}
obj <- QCFilter(obj, min_area = 20, max_area = 1000)
obj <- FindNeighbours(obj, k = 5)
obj <- CellDensity(obj, radius = 40)
a <- obj[obj$sample_id == "tonsil_A", ]
head(PhenotypeSummary(obj), 4)
```

```{r cellmap, dev = "ragg_png", fig.height = 4.8, fig.cap = "Phenotype map of one tonsil section."}
CellMap(a)
```

# A coherent result API

Every statistic returns a classed object with a tidy print and a one-line
`autoplot()`, while still behaving like its underlying data frame.

```{r results}
rk <- RipleysK(a, r_seq = seq(0, 120, length.out = 40), correction = "border")
rk
```

```{r autoplot, fig.height = 3.4, fig.cap = "autoplot() of Ripley's L and the neighbourhood-enrichment z-scores."}
autoplot(rk)
autoplot(NeighbourhoodEnrichment(a, radius = 40, n_perm = 199, seed = 1))
```

# Rigorous statistics

```{r rigorous}
# Translation edge correction
head(RipleysK(a, r_seq = seq(0, 100, length.out = 20),
              correction = "translation")$K, 4)

# Moran's I with inverse-distance weights + permutation p-value
MoransI(a, feature = "CD20", radius = 40, weights = "idw", n_perm = 199,
        seed = 1)

# Interaction matrix with a permutation null
head(InteractionMatrix(a, radius = 40, method = "permutation", n_perm = 99,
                       seed = 1), 4)
```

```{r scree, fig.height = 3.4, fig.cap = "PCA variance explained (scree plot)."}
obj <- RunPCA(obj, n_pcs = 8)
ScreePlot(obj)
```

\newpage

# Higher-order spatial structure

## Cellular neighbourhoods (niches)

```{r niches, dev = "ragg_png", fig.height = 4.8, fig.cap = "Cellular neighbourhoods: cells coloured by the niche their local composition belongs to."}
obj <- CellularNeighbourhoods(obj, n_neighbourhoods = 6, k = 20, seed = 1)
print(round(obj@spatial$neighbourhood_composition, 2))
CellMap(obj[obj$sample_id == "tonsil_A", ], colour_by = "neighbourhood")
```

## Spatial domains

```{r domains, dev = "ragg_png", fig.height = 4.8, fig.cap = "Spatial domains from smoothed expression."}
obj <- SpatialDomains(obj, n_domains = 4, k = 20, seed = 1)
CellMap(obj[obj$sample_id == "tonsil_A", ], colour_by = "domain")
```

# Unsupervised phenotyping

```{r gmm, eval = has_mclust}
obj <- ExpressionClusters(obj, k = 6, method = "gmm")
AnnotateClusters(obj, add_column = FALSE)
```

```{r gmm-skip, echo = FALSE, results = "asis", eval = !has_mclust}
cat("_mclust not installed; GMM phenotyping skipped._\n")
```

# Differential abundance

```{r diffabund}
obj@meta_data$arm <- ifelse(obj$sample_id == "tonsil_A", "ctrl", "treat")
DifferentialAbundance(obj, condition = "arm")
```

\newpage

# Platform readers

A general expression-plus-coordinates reader underlies per-vendor wrappers for
Xenium, CosMx, and MERSCOPE.

```{r readers}
expr <- matrix(rpois(40, 5), nrow = 10,
               dimnames = list(NULL, c("CD3", "CD8", "CD20", "PanCK")))
meta <- data.frame(x = runif(10, 0, 100), y = runif(10, 0, 100))
ReadMatrixCoords(expr, meta)
```

# Ecosystem interoperability

```{r interop, eval = has_seurat}
se <- as_Seurat(phenoscapR_example)
class(se)
as_SpatialCellData(se)
```

```{r interop-skip, echo = FALSE, results = "asis", eval = !has_seurat}
cat("_Seurat not installed; interop demonstration skipped._\n")
```

# Performance

All neighbour-based statistics route through a kd-tree search engine (RANN,
with an exact base-R fallback) and memory-bounded streaming for the
radius-accumulating statistics. A 500,000-cell, ten-section cohort builds in
~2 s, and each whole-cohort statistic (density, neighbours, interactions)
completes in seconds rather than requiring a multi-gigabyte distance matrix.
See the companion "phenoscapR at Scale" report for the full benchmark.

# Session information

```{r session}
sessionInfo()
```
