---
title: "phenoscapR — Function Showcase"
subtitle: "Single-cell spatial biology, end to end, on the bundled example data"
author: "R. Heller"
date: "`r format(Sys.Date())`"
output:
  pdf_document:
    toc: true
    toc_depth: 2
    number_sections: true
    fig_caption: true
    df_print: kable
fontsize: 10pt
geometry: margin=2.2cm
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>", echo = TRUE,
  message = FALSE, warning = FALSE,
  fig.width = 6.5, fig.height = 4, fig.align = "center", dpi = 150
)
options(width = 90)
has_deldir <- requireNamespace("deldir", quietly = TRUE)
has_uwot   <- requireNamespace("uwot", quietly = TRUE)
```

\newpage

# Overview

**phenoscapR** takes multiplexed-imaging cell-segmentation data from raw table
to publication-ready spatial statistics and figures. This document runs the
*entire* public API on the bundled synthetic dataset `phenoscapR_example`
(two tonsil sections, eight markers, planted spatial niches), so every figure
below is real output, reproducible with the code shown.

```{r load}
library(phenoscapR)
library(ggplot2)
data(phenoscapR_example)
phenoscapR_example
```

The data ship with a ground-truth population per cell:

```{r truth}
table(phenoscapR_example$sample_id, phenoscapR_example$phenotype_true)
```

# The processing pipeline

## Quality control — `QCFilter()`

```{r qc}
obj <- QCFilter(phenoscapR_example, min_area = 20, max_area = 1000)
obj
```

## Normalisation — `NormaliseData()`

```{r norm}
obj <- NormaliseData(obj, method = "zscore")
round(GetData(obj)[1:4, ], 2)
```

## Phenotyping — `PhenotypeCells()` / `PhenotypeSummary()`

```{r pheno}
obj <- PhenotypeCells(obj, thresholds = list(
  CD20 = 1, CD3 = 1, CD8 = 1, CD4 = 1, CD68 = 1, PanCK = 1, FoxP3 = 1))
head(PhenotypeSummary(obj), 8)
```

For readable figures we use the curated ground-truth populations from here on:

```{r usetruth}
obj@meta_data$phenotype <- obj@meta_data$phenotype_true
obj <- CellDensity(obj, radius = 40)
obj <- FindNeighbours(obj, k = 5)
a <- obj[obj$sample_id == "tonsil_A", ]   # a single tissue section
markers <- c("CD20", "CD3", "CD8", "PanCK")
```

\newpage

# Tissue maps

## `CellMap()` — phenotype map

```{r cellmap, fig.height = 4.6, fig.cap = "Every cell at its tissue position, coloured by phenotype. The B-cell follicle and T-cell zone are clearly resolved."}
CellMap(a)
```

## `FeaturePlot()` — marker intensity in space

```{r featureplot, fig.height = 5, fig.cap = "Normalised marker intensity per cell."}
FeaturePlot(a, features = markers)
```

## `DensityPlot()` — local cell density

```{r density, fig.height = 4.6, fig.cap = "Local neighbour count within a 40-unit radius."}
DensityPlot(a)
```

\newpage

# Marker distributions

## `ViolinPlot()` and `BoxPlot()`

```{r violin, fig.height = 4, fig.cap = "Per-phenotype marker distributions."}
ViolinPlot(a, features = markers, group_by = "phenotype")
```

```{r box, fig.height = 4, fig.cap = "Box-plot view of the same distributions."}
BoxPlot(a, features = markers, group_by = "phenotype")
```

## `RidgePlot()`, `HistogramPlot()` and `DotPlot()`

```{r ridge, fig.height = 4.2, fig.cap = "Ridge plot of marker intensities by phenotype."}
RidgePlot(a, features = markers, group_by = "phenotype")
```

```{r hist, fig.height = 3.6, fig.cap = "CD20 intensity histogram, split by phenotype."}
HistogramPlot(a, feature = "CD20", group_by = "phenotype")
```

```{r dot, fig.height = 3.8, fig.cap = "Mean expression (colour) and fraction expressing (size) per phenotype."}
DotPlot(a, features = c("CD3", "CD4", "CD8", "CD20", "CD68", "PanCK", "FoxP3"))
```

\newpage

# Summaries

## `MarkerHeatmap()` — phenotype signatures

```{r heatmap, fig.height = 4.6, fig.cap = "Mean marker expression per phenotype."}
MarkerHeatmap(obj)
```

## `CompositionPlot()` — phenotype proportions

```{r composition, fig.height = 3.6, fig.cap = "Phenotype composition per sample."}
CompositionPlot(obj, group_by = "sample_id")
```

## Phenotype interactions — `InteractionMatrix()` / `InteractionPlot()`

`InteractionMatrix()` counts neighbour pairs per sample (never across tissues)
and scores them as `log2(observed / expected)`.

```{r interactions, fig.height = 4.4, fig.cap = "Spatial interaction scores between phenotypes."}
im <- InteractionMatrix(obj, radius = 40)
head(im[order(-im$interaction_score), ], 6)
InteractionPlot(im)
```

\newpage

# Point-pattern statistics

These describe a single observation window, so they run on one section `a`.

## `RipleysK()` — multi-scale clustering

```{r ripley, fig.height = 3.8, fig.cap = "Border-corrected Ripley's L. Values above zero indicate clustering."}
rk <- RipleysK(a, r_seq = seq(0, 120, length.out = 40), correction = "border")
with(rk, { plot(r, L, type = "l", lwd = 2, col = "#3366cc",
  xlab = "r", ylab = "L(r)", main = "Ripley's L (border-corrected)")
  abline(h = 0, lty = 2) })
```

## `PairCorrelation()` — clustering at a given distance

```{r pcf, fig.height = 3.8, fig.cap = "Pair correlation g(r); >1 means clustering at distance r."}
pcf <- PairCorrelation(a, r_seq = seq(5, 120, length.out = 40))
with(pcf, { plot(r, g, type = "l", lwd = 2, col = "#cc3366",
  xlab = "r", ylab = "g(r)", main = "Pair correlation function")
  abline(h = 1, lty = 2) })
```

## `MoransI()`, `QuadratAnalysis()`, `NeighbourhoodEnrichment()`

```{r stats}
# Spatial autocorrelation of a lineage marker
unlist(MoransI(a, feature = "CD20", radius = 50))

# Chi-squared test of spatial randomness
qa <- QuadratAnalysis(a, nx = 5, ny = 5)
qa[c("chi_sq", "p_value", "VMR")]

# Permutation test of phenotype co-localisation (top pairs)
ne <- NeighbourhoodEnrichment(a, radius = 40, n_perm = 199, seed = 1)
head(ne[order(-ne$z_score), ], 6)
```

## `CrossNNDistance()` — directed proximity

```{r crossnn}
summary(CrossNNDistance(a, from = "T helper", to = "B cell"))
```

## `DelaunayNetwork()` / `SpatialNetworkPlot()` — contact graph

```{r delaunay, eval = has_deldir, fig.height = 4.6, fig.cap = "Delaunay cell-contact graph (edges capped at 60 units)."}
a_net <- DelaunayNetwork(a, max_edge = 60)
SpatialNetworkPlot(a_net)
```

\newpage

# Dimensionality reduction

## `RunPCA()` + `DimPlot()`

```{r pca, fig.height = 4.4, fig.cap = "PCA of marker space, coloured by phenotype."}
obj <- RunPCA(obj, n_pcs = 10)
DimPlot(obj, reduction = "pca", colour_by = "phenotype")
```

## `RunUMAP()` + `EmbeddingPlot()`

```{r umap, eval = has_uwot, fig.height = 4.4, fig.cap = "UMAP embedding of marker space."}
obj <- RunUMAP(obj, dims = 10, n_neighbors = 15)
EmbeddingPlot(obj, reduction = "umap", colour_by = "phenotype")
```

# Palettes

The global palette drives every plot; `SetPalette()` accepts any viridis-family
name or a custom gradient.

```{r palette, fig.height = 3.4, fig.cap = "The same composition plot under the 'rocket' palette."}
SetPalette("rocket")
CompositionPlot(obj, group_by = "sample_id")
```

```{r reset, include = FALSE}
SetPalette("viridis")
```

# Session information

```{r session}
sessionInfo()
```
