---
title: "Analysis"
output:
  workflowr::wflow_html:
    toc: true
editor_options:
  chunk_output_type: console
---

# Packages

```{r}
# if (!require("BiocManager", quietly = TRUE))
#     install.packages("BiocManager")

# BiocManager::install("SpatialExperiment")
# BiocManager::install("DropletUtils")
# BiocManager::install("BumpyMatrix")
# BiocManager::install("hoodscanR")

library(tidyverse)

# install.packages('Seurat')
library(Seurat)

library(Seurat)
# library(future); plan("multisession", workers = 24)
library(dplyr)
library(ggplot2)
library(patchwork)  # for combined ggplots

library(pbapply)



```


# Load rds

```{r}
codex.obj.lst <- readRDS("./data/df.rds")

# names(codex.obj.lst)
# 
# tFL1 <- codex.obj.lst$AnnotFL1
# tFL2 <- codex.obj.lst$tFL2Anno
# tFL3 <- codex.obj.lst$'tFL3 Anno'
# tFL4 <- codex.obj.lst$AnnotFL4
# tFL5 <- codex.obj.lst$AnnoTfl5
# tFL6 <- codex.obj.lst$'AnnoTfl 6'
# tFL7 <- codex.obj.lst$'Anno tFL7'
# tFL8 <- codex.obj.lst$AnnotFL8
# 
# 
# # Define groups
# FL_samples <- list(tFL6, tFL3, tFL7, tFL8)
# DLBCL_samples <- list(tFL1, tFL4, tFL2, tFL5)
# all_samples <- c(FL_samples, DLBCL_samples)
# names(all_samples) <- c("FL6", "FL3", "FL7", "FL8", "FL1", "FL4", "FL2", "FL5")

# sample_ids <- c("tFL6","tFL3","tFL7","tFL8","tFL1","tFL4","tFL2","tFL5")


# Get cell counts per processed sample
sample_sizes <- sapply(codex.obj.lst, function(obj) ncol(obj))
min_cells <- min(sample_sizes)
max_cells <- max(sample_sizes)
max_cells/min_cells


# Downsample Each Sample to min_cells
set.seed(123)  # for reproducibility
codex.obj.lst_downsample <- lapply(codex.obj.lst, function(obj) {
  cells_to_keep <- sample(colnames(obj), size = min_cells, replace = FALSE)
  subset(obj, cells = cells_to_keep)
})



```


# Formulary

```{r}
f_process_sample <- function(obj) {
  # DefaultAssay(obj) <- "Akoya"
  obj <- NormalizeData(obj, normalization.method = "CLR", margin = 2) # margin == normalize across features (1) or cells (2)
  obj <- ScaleData(obj, features = rownames(obj))
  return(obj)
}

```



# Flow




## Preprocess & Z‑Standardise All Samples

```{r}
# processed_list <- lapply(all_samples, f_process_sample)
codex.obj.lst_p <- pblapply(codex.obj.lst_downsample, f_process_sample)

```

## Merge Samples While Preserving IDs

```{r}
# Prepare lists
sample_ids <- names(codex.obj.lst_p)

# Merge all at once
codex.obj_all <- merge(
  x = codex.obj.lst_p[[1]],
  y = codex.obj.lst_p[-1],
  add.cell.ids = sample_ids,
  project = "combined",
  merge.data = TRUE,
  merge.dr = FALSE
)

str(codex.obj_all)
print(codex.obj_all)
```



## 🔄 1. Collapse Layered Assays into a Single Matrix

Your merged Seurat object currently has separate `counts.1`, `data.1`, ..., `scale.data.8` layers for each sample. To operate seamlessly, collapse these into unified assay layers:

```{r}

DefaultAssay(codex.obj_all) <- "Akoya"
codex.obj_all[["Akoya"]] <- JoinLayers(codex.obj_all[["Akoya"]])
```

This consolidates all raw counts, normalized data, and scaled data into single-layer sets—ensuring later functions use pooled data ([biostars.org][1]).


## 🔬 2. Dimensional Reduction & Clustering

```{r}
# Switch to protein assay
DefaultAssay(codex.obj_all) <- "Akoya"

# Identify most variable markers
codex.obj_all <- FindVariableFeatures(codex.obj_all, 
                                      selection.method = "vst", 
                                      nfeatures = nrow(GetAssayData(codex.obj_all)))

# Run PCA
codex.obj_all <- RunPCA(codex.obj_all, 
                        features = VariableFeatures(codex.obj_all), 
                        npcs = 8)

# Find out how many PCs were computed
num_pcs <- ncol(codex.obj_all[["pca"]]@cell.embeddings)

# Build neighbor graph & cluster
codex.obj_all <- FindNeighbors(codex.obj_all, 
                               reduction = "pca", 
                               dims = 1:num_pcs)
codex.obj_all <- FindClusters(codex.obj_all, resolution = 0.2)

# Compute UMAP & t-SNE
codex.obj_all <- RunUMAP(codex.obj_all, reduction = "pca", dims = 1:num_pcs)
codex.obj_all <- RunTSNE(codex.obj_all, reduction = "pca", dims = 1:num_pcs)

print(codex.obj_all)
str(codex.obj_all)
```


```{r}

```



## Assign groups

```{r}

# Example: assign FL vs DLBCL based on sample
FL_ids <- c("tFL6","tFL3","tFL7","tFL8")
codex.obj_all$sample <- sapply(strsplit(colnames(codex.obj_all), "_"), `[`, 1)
codex.obj_all$clinical_group <- ifelse(codex.obj_all$sample %in% FL_ids, "FL", "DLBCL")

Idents(codex.obj_all) <- codex.obj_all$seurat_clusters

markers_groups <- FindAllMarkers(
  object = codex.obj_all,
  assay = "Akoya",
  only.pos = TRUE,
  group.by = "clinical_group",
  logfc.threshold = 0.25,
  min.pct = 0.1
)

markers_FL_vs_DLBCL <- FindMarkers(
  object = codex.obj_all,
  assay = "Akoya",
  group.by = "clinical_group",
  ident.1 = "FL",
  ident.2 = "DLBCL",
  test.use = "wilcox",
  logfc.threshold = 0.25,
  min.pct = 0.1
)


markers <- FindAllMarkers(codex.obj_all, assay = "Akoya", only.pos = TRUE)



```


```{r}
library(dplyr)

markers <- markers_FL_vs_DLBCL %>%
  rownames_to_column("feature") %>%
  mutate(
    log10_padj = -log10(p_val_adj + 1e-300) * sign(avg_log2FC),
    threshold = (abs(avg_log2FC) > 1 & p_val_adj < 1e-5 & pct.1 > 0.1)
  )

library(ggplot2)

ggplot(markers, aes(x = avg_log2FC, y = log10_padj)) +
  geom_point(aes(color = threshold, size = pct.1 - pct.2), alpha = 0.7) +
  scale_color_manual(values = c("grey70", "red")) +
  geom_vline(xintercept = c(-1, 1), linetype = "dashed", color = "blue") +
  geom_hline(yintercept = -log10(1e-5), linetype = "dashed", color = "blue") +
  theme_minimal() +
  labs(
    x = "Avg log2 Fold Change (FL vs DLBCL)",
    y = "-log10(Adjusted p-value) * sign(FC)",
    color = "Significant",
    size = "Pct.1 – Pct.2"
  ) +
  theme(legend.position = "right")

markers <- markers %>% 
  mutate(mean_expr = (pct.1 + pct.2) / 2)

ggplot(markers, aes(x = mean_expr, y = avg_log2FC)) +
  geom_point(aes(color = threshold), alpha = 0.7) +
  scale_color_manual(values = c("grey70", "red")) +
  geom_hline(yintercept = 0, color = "black") +
  theme_minimal() +
  labs(
    x = "Average Detection Rate (Pct)",
    y = "Avg log2 Fold Change",
    color = "Significant"
  ) +
  theme(legend.position = "right")


library(ggrepel)

top_hits <- markers %>% filter(threshold) %>% arrange(p_val_adj) %>% head(10)

ggplot(markers, aes(avg_log2FC, log10_padj)) +
  geom_point(aes(color = threshold), alpha = 0.7) +
  geom_text_repel(data = top_hits, aes(label = feature), size = 3) +
  scale_color_manual(values = c("grey70", "red")) +
  theme_minimal()

```




## 🏷️ 3. Manual Cluster Annotation

Identify marker proteins for each cluster:

```{r}
markers <- FindAllMarkers(codex.obj_all, assay = "Akoya", only.pos = TRUE)
top5 <- markers %>% group_by(cluster) %>% slice_max(avg_log2FC, n = 5)
print(top5)

# write markers to xlsx
writexl::write_xlsx(markers, "markers.xlsx")
writexl::write_xlsx(markers_FL_vs_DLBCL, "markers_FL_vs_DLBCL.xlsx")




```

Then manually map clusters to cell types:

```{r}
map_clusters_to_celltypes <- function(cl) {
  switch(as.character(cl),
    "0" = "B cells",
    "1" = "T helper cells",
    "2" = "proliferative B cells",
    "3" = "Cytotoxic T cells",
    "4" = "Myeloid cells",
    "5" = "ECM",
    "6" = "immune cell infiltrate",
    "Unknown")
}
codex.obj_all$manual_label <- sapply(codex.obj_all$seurat_clusters, map_clusters_to_celltypes)


```



## 🌐 4. Visualization

* **UMAP** colored by cell types:

```{r}
  DimPlot(codex.obj_all, reduction = "umap", group.by = "manual_label", label = TRUE)
```

* **Split by clinical group or sample**:

```{r}
  DimPlot(codex.obj_all, reduction = "umap", group.by = "manual_label", split.by = "clinical_group")
```

```{r}
# UMAP colored by manual cell-type labels
DimPlot(codex.obj_all, reduction = "umap",
        group.by = "manual_label",
        split.by = "clinical_group",
        label = TRUE, repel = TRUE) +
  ggtitle("Cell-types by Clinical Group")

# Ki67 expression overlay
FeaturePlot(codex.obj_all, features = "Ki67..Cell..Mean",
            reduction = "umap",
            split.by = "clinical_group",
            cols = c("lightgrey", "red"))

```

```{r}
library(Seurat)

# Base UMAP plot
p <- DimPlot(
  codex.obj_all,
  reduction = "umap",
  group.by = "manual_label",
  split.by = "clinical_group",
  # label = TRUE,
  repel = TRUE,
  pt.size = 2,       # increase point size
  alpha = 0.6        # semi-transparent points
)

# Apply a dark background and remove default title
p + DarkTheme() +    # black background, white text :contentReference[oaicite:1]{index=1}
    theme(plot.title = element_blank())  # removes the plot title

```

```{r}
library(Seurat)
library(patchwork)

# Base FeaturePlot
p_f <- FeaturePlot(
  object    = codex.obj_all,
  features  = "Ki67..Cell..Mean",
  reduction = "umap",
  split.by  = "clinical_group",
  cols      = c("lightgrey", "red"),
  pt.size   = .2,
  order     = TRUE,
  combine = TRUE  # keep separate for patchwork
) +
  scale_color_gradient(low = "lightgrey", high = "red") +
  labs(color = "Ki67") +   # rename legend title
  theme_void() +            # minimal plot background
  theme(
    legend.position = "right",
    strip.text = element_text(color = "white", size = 14),
    legend.title = element_text(color = "white"),
    legend.text = element_text(color = "white")
  )

# Add dark backgrounds to each panel
dark_panels <- theme(
  panel.background = element_rect(fill = "black", color = NA),
  plot.background  = element_rect(fill = "black", color = NA)
)

# Apply across the patchwork layout
(p_f + plot_layout(ncol = 2)) &
  dark_panels


```

```{r}
library(Seurat)
library(patchwork)
library(ggplot2)

# 1. Reorder clinical_group factor to ensure FL is first
codex.obj_all$clinical_group <- factor(
  codex.obj_all$clinical_group,
  levels = c("FL", "DLBCL")
)

# 2. Create base FeaturePlot
p_f <- FeaturePlot(
  object    = codex.obj_all,
  features  = "Bcl-2..Cell..Mean",
  reduction = "umap",
  split.by  = "clinical_group",
  cols      = c("lightgrey", "red"),
  pt.size   = .2,
  order     = TRUE
) +
  scale_color_gradient(low = "lightgrey", high = "red") +
  labs(color = "Bcl-2") +
  theme_void() +
  theme(
    legend.position = "right",
    strip.text = element_text(color = "white", size = 14),
    legend.title = element_text(color = "white"),
    legend.text = element_text(color = "white")
  )

# 3. Define dark theme elements
dark_layers <- theme(
  panel.background = element_rect(fill = "black", color = NA),
  plot.background  = element_rect(fill = "black", color = NA),
  strip.background = element_rect(fill = "black", color = NA)
)

# 4. Apply dark theme across the patchwork
final_plot <- (p_f + plot_layout(ncol = 2)) & dark_layers

# 5. Render the plot
final_plot

```




```{r}
library(Seurat)
library(patchwork)

# Ensure t-SNE is computed; if not, run:
if ( !"tsne" %in% names(codex.obj_all@reductions) ) {
  codex.obj_all <- RunTSNE(codex.obj_all, reduction = "pca", dims = 1:num_pcs)
}

# 1️⃣ t-SNE colored by manual annotations
p1 <- DimPlot(
  codex.obj_all,
  reduction = "tsne",
  group.by = "manual_label",
  split.by = "clinical_group",
  label = TRUE,
  repel = TRUE
) + ggtitle("t‑SNE: Manual Cell Types")

# 2️⃣ t-SNE showing Ki67 expression
p2 <- FeaturePlot(
  codex.obj_all,
  features = "Ki67..Cell..Mean",
  reduction = "tsne",
  split.by = "clinical_group",
  cols = c("lightgrey", "red"),
  pt.size = 0.5
) + ggtitle("t‑SNE: Ki67 Expression")

# Combine with patchwork
(p1 | p2) + plot_layout(guides = "collect")

```


```{r}
library(Seurat)
library(patchwork)

# Ensure UMAP is computed; if not, run (using your PCA dims):
if (! "umap" %in% names(codex.obj_all@reductions)) {
  codex.obj_all <- RunUMAP(codex.obj_all, reduction = "pca", dims = 1:num_pcs)
}

# 1️⃣ UMAP styled by manual annotations
p1 <- DimPlot(
  codex.obj_all,
  reduction = "umap",
  group.by = "manual_label",
  split.by = "clinical_group",
  label = TRUE,
  repel = TRUE
) + ggtitle("UMAP: Manual Cell Types")

# 2️⃣ UMAP displaying Ki67 expression
p2 <- FeaturePlot(
  codex.obj_all,
  features = "Ki67..Cell..Mean",
  reduction = "umap",
  split.by = "clinical_group",
  cols = c("lightgrey", "red"),
  pt.size = 0.5
) + ggtitle("UMAP: Ki67 Expression")

# Combine plots
p1 | p2 + plot_layout(guides = "collect")

```



* **Spatial distributions** (if spatial metadata/images are available):

```{r}
  SpatialDimPlot(codex.obj_all, group.by = "manual_label")
```


## 📊 5. Compare Cell Type Frequencies Across Groups

```{r}
df <- codex.obj_all@meta.data %>%
  count(clinical_group, manual_label) %>%
  group_by(clinical_group) %>%
  mutate(pct = n / sum(n) * 100)
ggplot(df, aes(clinical_group, pct, fill = manual_label)) +
  geom_col(position = "fill") +
  labs(y = "Proportion (%)")
```

Run a statistical test:

```{r}
chisq.test(table(codex.obj_all$clinical_group, codex.obj_all$manual_label))
```




```{r}
# 1. Subset the combined Seurat object for tFL6 cells
tFL6 <- subset(codex.obj_all, subset = sample == "tFL6")

ImageDimPlot(
  object     = tFL6,
  group.by   = "manual_label",
  size = 3,
  alpha = 0.5

) + ggtitle("tFL6 – Manual Cell-Type Spatial Plot")

```





```{r}
obj <- tFL6

cluster_markers <- FindAllMarkers(obj)
top_markers <- cluster_markers %>%
  group_by(cluster) %>%
  slice_max(order_by = avg_log2FC, n = 5)
print(top_markers)


# Standardize, cluster, annotate (manual annotation placeholder)
process_obj <- function(obj) {
  obj <- NormalizeData(obj, normalization.method = "CLR")
  obj <- ScaleData(obj, features = rownames(obj))
  obj <- FindVariableFeatures(obj)
  obj <- RunPCA(obj, features = VariableFeatures(obj))
  num_pcs <- ncol(obj[["pca"]]@cell.embeddings)
  obj <- FindNeighbors(obj, dims = 1:num_pcs)
  obj <- FindClusters(obj, resolution = 0.5, future.seed=TRUE)
  obj$auto_cluster <- Idents(obj)
  # Add manual label column here, e.g.:
  obj$manual_label <- map_clusters_to_celltypes(obj$auto_cluster)
  return(obj)
}

FL_processed <- lapply(FL_samples, process_obj)
DLBCL_processed <- lapply(DLBCL_samples, process_obj)
all_samples <- c(FL_processed, DLBCL_processed)
names(all_samples) <- c(paste0("FL", 3:8), paste0("DLBCL", c(1,4,2,5)))

# 4. Run UMAP/t-SNE per sample
all_samples <- lapply(all_samples, function(obj) {
  obj <- RunUMAP(obj, dims = 1:20)
  obj <- RunTSNE(obj, dims = 1:20)
  return(obj)
})

# 5. Merge by clinical group
FL_merged <- merge(FL_processed[[1]], y = FL_processed[-1], add.cell.ids = names(FL_processed), project = "FL_group")
DLBCL_merged <- merge(DLBCL_processed[[1]], y = DLBCL_processed[-1], add.cell.ids = names(DLBCL_processed), project = "DLBCL_group")
both <- merge(FL_merged, y = DLBCL_merged, add.cell.ids = c("FL", "DLBCL"), project = "combined")

# 6. Re-process merged
both <- NormalizeData(both)
both <- ScaleData(both)
both <- FindVariableFeatures(both, selection.method = "vst", nfeatures = 2000)
both <- RunPCA(both, dims = 1:20)
both <- FindNeighbors(both, dims = 1:20)
both <- FindClusters(both, resolution = 0.5)
both <- RunUMAP(both, dims = 1:20)
both <- RunTSNE(both, dims = 1:20)
Idents(both) <- both$seurat_clusters

# 7. Add metadata for clinical_group (FL vs DLBCL)
both$clinical_group <- ifelse(grepl("^FL", both$orig.ident), "FL", "DLBCL")

# 8. Visualizations
p_umap <- DimPlot(both, reduction = "umap", group.by = "clinical_group", label = TRUE) + ggtitle("UMAP by clinical group")
p_tsne <- DimPlot(both, reduction = "tsne", group.by = "clinical_group", label = TRUE) + ggtitle("t-SNE by clinical group")

# Optional: colored by manual_label if available
# p_umap_label <- DimPlot(both, reduction = "umap", group.by = "manual_label", label = TRUE)

# Spatial layout (if spatial coords exist per sample):
# lapply(all_samples, function(obj) SpatialDimPlot(obj, label = TRUE))

# 9. Proportional barplot of cell types in each group
df <- both@meta.data %>%
  count(clinical_group, auto_cluster) %>%
  group_by(clinical_group) %>%
  mutate(percent = n / sum(n) * 100)

p_bar <- ggplot(df, aes(x = clinical_group, y = percent, fill = auto_cluster)) +
  geom_bar(stat = "identity", position = "fill") +
  labs(y = "Proportion of cell clusters") +
  ggtitle("Cell cluster distribution by clinical group")

# 10. Statistical test
tbl <- table(both$clinical_group, both$auto_cluster)
chi_res <- chisq.test(tbl)

# 11. Display
print(p_umap + p_tsne + p_bar)
print(chi_res)
```


# Analysis

## tFL1

```{r}
codex.obj <- NormalizeData(object = tFL6, normalization.method = "CLR", margin = 2)
codex.obj <- ScaleData(codex.obj)
VariableFeatures(codex.obj) <- rownames(codex.obj)  # since the panel is small, treat all features as variable.
codex.obj <- RunPCA(object = codex.obj, npcs = 20, verbose = FALSE)
codex.obj <- RunUMAP(object = codex.obj, dims = 1:10, verbose = FALSE)
codex.obj <- FindNeighbors(object = codex.obj, dims = 1:10, verbose = FALSE)
codex.obj <- FindClusters(object = codex.obj, verbose = FALSE, resolution = 0.4, n.start = 1)
```






Unsupervised analysis to identify cell clusters.


```{r}
codex.obj <- NormalizeData(object = codex.obj, normalization.method = "CLR", margin = 2)
codex.obj <- ScaleData(codex.obj)
VariableFeatures(codex.obj) <- rownames(codex.obj)  # since the panel is small, treat all features as variable.
codex.obj <- RunPCA(object = codex.obj, npcs = 20, verbose = FALSE)
codex.obj <- RunUMAP(object = codex.obj, dims = 1:20, verbose = FALSE)
codex.obj <- FindNeighbors(object = codex.obj, dims = 1:20, verbose = FALSE)
codex.obj <- FindClusters(object = codex.obj, verbose = FALSE, resolution = 0.4, n.start = 1)
DimPlot(codex.obj, label = TRUE, label.box = TRUE)

```


# Visualisation

Visualize the cell clusters based on a protein intensity-based UMAP embedding, or also based on their spatial location.

## UMAP

```{r}
DimPlot(codex.obj, label = TRUE, label.box = TRUE)
```

# Histo overlay / spatial location

```{r}
ImageDimPlot(codex.obj, cols = "parade")
```

The expression patters of individual markers clearly denote different cell types and spatial structures, including Lyve1 (lymphatic endothelial cells), CD34 (blood endothelial cells), and CD21 (B cells). As expected, endothelial cells group together into vessels, and B cells are key components of specialized microstructures known as germinal zones.

```{r}
# p1 <- ImageFeaturePlot(codex.obj, fov = "Test", features = c("CD34", "CD21", "Lyve1"),
#     min.cutoff = "q10", max.cutoff = "q90")
# p2 <- ImageDimPlot(codex.obj, fov = "Test", cols = "parade")
# p1 + p2
```
