---
title: "Advanced Capabilities"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Advanced Capabilities}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<!-- badges: start -->
[![R-CMD-check](https://github.com/CTTIR/phenoscapR/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/CTTIR/phenoscapR/actions/workflows/R-CMD-check.yaml)
[![pkgdown](https://github.com/CTTIR/phenoscapR/actions/workflows/pkgdown.yaml/badge.svg)](https://cttir.github.io/phenoscapR/)
[![CRAN status](https://www.r-pkg.org/badges/version/phenoscapR)](https://CRAN.R-project.org/package=phenoscapR)
[![Codecov test coverage](https://codecov.io/gh/CTTIR/phenoscapR/branch/main/graph/badge.svg)](https://app.codecov.io/gh/CTTIR/phenoscapR?branch=main)
[![CRAN downloads](https://cranlogs.r-pkg.org/badges/phenoscapR)](https://cran.r-project.org/package=phenoscapR)
[![CRAN downloads total](https://cranlogs.r-pkg.org/badges/grand-total/phenoscapR)](https://cran.r-project.org/package=phenoscapR)
[![License: MIT](https://img.shields.io/badge/license-MIT-yellow.svg)](https://opensource.org/licenses/MIT)
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
<!-- badges: end -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  fig.width = 7, fig.height = 4.4, dpi = 96
)
has_mclust <- requireNamespace("mclust", quietly = TRUE)
has_seurat <- requireNamespace("Seurat", quietly = TRUE) &&
  requireNamespace("SeuratObject", quietly = TRUE)
```

<img src="../man/figures/logo.svg" align="right" height="139" alt="phenoscapR hex logo" style="margin-left:12px;margin-bottom:4px;"/>

This vignette tours the capabilities added beyond the core workflow: a coherent
result API, higher-order spatial structure, rigorous statistics, unsupervised
phenotyping, differential testing, platform readers, and ecosystem interop.

```{r data}
library(phenoscapR)
library(ggplot2)
data(phenoscapR_example)

obj <- phenoscapR_example
obj@meta_data$phenotype <- obj@meta_data$phenotype_true
obj <- NormaliseData(obj, "zscore")
a <- obj[obj$sample_id == "tonsil_A", ]
```

## A coherent result API

Every spatial statistic returns a classed object with a tidy `print` and a
one-line `autoplot()`, while still behaving like its underlying data frame.

```{r results}
rk <- RipleysK(a, r_seq = seq(0, 120, length.out = 40), correction = "border")
rk                       # tidy print
head(rk$L)               # ...but still a data frame
```

```{r autoplot, fig.height = 3.6}
autoplot(rk)
autoplot(NeighbourhoodEnrichment(a, radius = 40, n_perm = 199, seed = 1))
```

### Per-sample analysis

The single-window statistics refuse multi-sample objects by default, but
`by_sample = TRUE` maps them across every section and returns a named list.

```{r bysample}
per <- RipleysK(obj, r_seq = seq(0, 100, length.out = 20), by_sample = TRUE)
per
per[["tonsil_B"]]
```

## Rigorous statistics

Ripley's K offers a translation edge correction; Moran's I offers weight
schemes and a permutation p-value; the interaction matrix offers a permutation
null.

```{r rigorous}
RipleysK(a, r_seq = seq(0, 100, length.out = 20),
         correction = "translation")$K |> head()

MoransI(a, feature = "CD20", radius = 40, weights = "idw", n_perm = 199,
        seed = 1)

InteractionMatrix(a, radius = 40, method = "permutation", n_perm = 99,
                  seed = 1) |> head(4)
```

### PCA variance

```{r pca, fig.height = 3.6}
obj <- RunPCA(obj, n_pcs = 8)
round(VarianceExplained(obj), 1)
ScreePlot(obj)
```

## Higher-order spatial structure

### Cellular neighbourhoods (niches)

Cluster cells by the phenotype composition of their local neighbourhood.

```{r niches, fig.height = 4.6}
obj <- CellularNeighbourhoods(obj, n_neighbourhoods = 6, k = 20, seed = 1)
table(obj$neighbourhood)
CellMap(obj[obj$sample_id == "tonsil_A", ], colour_by = "neighbourhood")
```

### Spatial domains

Cluster cells by spatially smoothed expression into coherent tissue regions.

```{r domains, fig.height = 4.6}
obj <- SpatialDomains(obj, n_domains = 4, k = 20, seed = 1)
CellMap(obj[obj$sample_id == "tonsil_A", ], colour_by = "domain")
```

## Unsupervised phenotyping

`ExpressionClusters()` clusters cells by expression (k-means, hierarchical, or
a Gaussian mixture model); `AnnotateClusters()` names clusters by their top
markers.

```{r gmm, eval = has_mclust}
obj <- ExpressionClusters(obj, k = 6, method = "gmm")
AnnotateClusters(obj, add_column = FALSE)
```

```{r gmm-note, echo = FALSE, results = "asis", eval = !has_mclust}
cat("> The `mclust` package is not installed, so the GMM example is skipped.\n")
```

## Differential abundance

Compare phenotype proportions across conditions, with the sample as the unit
of replication.

```{r diffabund}
obj@meta_data$arm <- ifelse(obj$sample_id == "tonsil_A", "ctrl", "treat")
DifferentialAbundance(obj, condition = "arm")
```

## Platform readers

A general expression-plus-coordinates reader underlies per-vendor wrappers
(`ReadXenium()`, `ReadCosMx()`, `ReadMERSCOPE()`).

```{r readers}
expr <- matrix(rpois(40, 5), nrow = 10,
               dimnames = list(NULL, c("CD3", "CD8", "CD20", "PanCK")))
meta <- data.frame(x = runif(10, 0, 100), y = runif(10, 0, 100),
                   region = rep(c("a", "b"), 5))
ReadMatrixCoords(expr, meta)
```

## Ecosystem interop

Convert to and from Seurat and SpatialExperiment.

```{r interop, eval = has_seurat}
se <- as_Seurat(phenoscapR_example)
class(se)
back <- as_SpatialCellData(se)
back
```

```{r interop-note, echo = FALSE, results = "asis", eval = !has_seurat}
cat("> The `Seurat` package is not installed, so the interop example is",
    "skipped.\n")
```

## See also

- [Getting Started](phenoscapR.html) -- the core workflow
- [Advanced Spatial Analysis](phenoscapR-03-spatial-analysis.html) -- the statistics
- [Visualisation Gallery](phenoscapR-04-visualisation.html) -- every plot
