---
title: "The SpatialCellData Object"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The SpatialCellData Object}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4
)
```

<img src="../man/figures/logo.svg" align="right" height="139" alt="phenoscapR hex logo" style="margin-left:12px;margin-bottom:4px;"/>

## Overview

The **`SpatialCellData`** S4 class is the central data container in
phenoscapR. It stores all information about a spatial biology experiment in
a single object:

| Slot | Type | Contents |
|------|------|----------|
| `counts` | numeric matrix | Raw marker intensities (cells × markers) |
| `data` | numeric matrix | Normalised intensities (filled by `NormaliseData()`) |
| `coords` | data.frame | Spatial coordinates (`x`, `y` columns) |
| `meta_data` | data.frame | Per-cell metadata (phenotype, sample ID, QC flags, …) |
| `project` | character | Project / experiment name |
| `spatial` | list | Results from spatial analysis functions |

This vignette walks through constructing, inspecting, and subsetting a
`SpatialCellData` object.

## Construction

### From raw matrices

Use `CreateSpatialObject()` when you already have a counts matrix and a
coordinates table:

```{r create}
library(phenoscapR)

set.seed(1)
n <- 200

counts <- matrix(
  abs(rnorm(n * 4, mean = 500, sd = 200)),
  nrow = n,
  dimnames = list(NULL, c("CD3", "CD8", "CD20", "PanCK"))
)

coords <- data.frame(
  x = runif(n, 0, 1000),
  y = runif(n, 0, 1000)
)

obj <- CreateSpatialObject(counts, coords, project = "demo")
obj
```

### From CSV files

Use `ReadSpatial()` to read one or more CSV files and return a
`SpatialCellData` object directly:

```{r read, eval = FALSE}
obj <- ReadSpatial("path/to/segmentation.csv", sample_id = "sample1")

# Or read a whole directory
obj <- ReadSpatial("path/to/csv_dir/")
```

## Inspecting the Object

```{r inspect}
# Dimensions: cells × markers
dim(obj)

# Cell count and marker names
NCells(obj)
NMarkers(obj)
Markers(obj)

# First few rows of metadata
head(Meta(obj))

# First few rows of coordinates
head(Coords(obj))
```

## Accessing Expression Data

```{r data}
# Raw counts
raw <- GetData(obj, slot = "counts")
raw[1:3, ]

# After normalisation, use slot = "data"
obj <- NormaliseData(obj, method = "zscore")
norm <- GetData(obj, slot = "data")
norm[1:3, ]
```

## Working with Metadata

Metadata columns are accessible with `$` and `[[`:

```{r meta}
# After phenotyping
obj <- PhenotypeCells(obj, thresholds = list(CD3 = 0, CD8 = 0,
                                              CD20 = 0, PanCK = 0))
head(obj$phenotype)
head(obj[["sample_id"]])
```

`Idents()` returns the active identity — phenotype labels if set,
otherwise sample identities:

```{r idents}
table(Idents(obj))
```

## Subsetting

`SpatialCellData` supports standard R subsetting with `[`:

```{r subset}
# Keep only CD3+ T cells
t_cells <- obj[obj$phenotype == "CD3+CD8-", ]
NCells(t_cells)
dim(t_cells)
```

## Next Steps

- **[Getting Started](phenoscapR.html)** — end-to-end workflow with the
  low-level data.table API
- **[Advanced Spatial Analysis](phenoscapR-03-spatial-analysis.html)** —
  Ripley's K, neighbourhood enrichment, Moran's I, and more
