---
title: "Getting Started with phenoscapR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with phenoscapR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 5
)
```

<img src="../man/figures/logo.svg" align="right" height="139" alt="phenoscapR hex logo" style="margin-left:12px;margin-bottom:4px;"/>

## Overview

**phenoscapR** provides tools for reading, processing, analysing, and
visualising single-cell spatial biology data from multiplexed imaging
platforms.

A typical workflow involves:

1. **Reading** cell segmentation data with `read_spatial()`
2. **Quality control** with `qc_filter()`
3. **Normalisation** with `normalise_markers()`
4. **Phenotyping** with `phenotype_cells()`
5. **Spatial analysis** with `nearest_neighbours()`, `cell_density()`,
   `interaction_matrix()`, and `spatial_clusters()`
6. **Visualisation** with `plot_cell_map()`, `plot_density()`,
   `plot_heatmap()`, and `plot_interactions()`

## Creating Example Data

Since real cell segmentation files can be large, we will create a small
simulated data set for illustration.

```{r simulate-data}
library(phenoscapR)

set.seed(42)
n_cells <- 500

# Simulate two cell populations in different spatial regions
dt <- data.table::data.table(
  `Cell ID` = seq_len(n_cells),
  `Cell X Position` = c(rnorm(250, 200, 50), rnorm(250, 600, 50)),
  `Cell Y Position` = c(rnorm(250, 300, 50), rnorm(250, 300, 50)),
  `Cell Area (px)` = rlnorm(n_cells, log(100), 0.3),
  CD3 = c(rnorm(250, 800, 150), rnorm(250, 200, 100)),
  CD8 = c(rnorm(250, 200, 100), rnorm(250, 700, 120)),
  DAPI = rnorm(n_cells, 1000, 200)
)

# Write to a temporary CSV
tmp <- tempfile(fileext = ".csv")
write.csv(dt, tmp, row.names = FALSE)
```

## Reading Data

```{r read-data}
cells <- read_spatial(tmp, sample_id = "example")
head(cells)
```

## Quality Control

```{r qc}
cells <- qc_filter(cells, min_area = 50, max_area = 500)
nrow(cells)
```

## Normalisation

```{r normalise}
cells <- normalise_markers(cells, method = "zscore")
summary(cells[, c("CD3", "CD8", "DAPI")])
```

## Phenotyping

```{r phenotype}
cells <- phenotype_cells(cells, thresholds = list(CD3 = 0, CD8 = 0))
summarise_phenotypes(cells)
```

## Spatial Analysis

### Nearest Neighbours

```{r nn}
cells <- nearest_neighbours(cells, k = 5)
summary(cells$nn_distance)
```

### Cell Density

```{r density}
cells <- cell_density(cells, radius = 50)
summary(cells$density)
```

### Interaction Matrix

```{r interactions}
interactions <- interaction_matrix(cells, radius = 50)
interactions
```

## Visualisation

### Cell Map

```{r cell-map, fig.cap="Cell phenotype map"}
plot_cell_map(cells, point_size = 1)
```

### Density Plot

```{r density-plot, fig.cap="Cell density map"}
plot_density(cells, point_size = 1)
```

### Marker Heatmap

```{r heatmap, fig.cap="Mean marker expression per phenotype"}
plot_heatmap(cells)
```

### Interaction Heatmap

```{r interaction-plot, fig.cap="Spatial interaction scores"}
plot_interactions(interactions)
```

```{r cleanup, include = FALSE}
unlink(tmp)
```
