---
title: "Getting Started with phenoscapR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with phenoscapR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 5
)
```

## Overview

**phenoscapR** provides an S4 class (`AkoyaExperiment`) and a pipe-friendly
workflow for analysing multiplexed spatial biology data from Akoya
Biosciences platforms (PhenoCycler, CODEX, PhenoImager) and QuPath.

The design follows the philosophy of the
[Seurat](https://satijalab.org/seurat/) package: a central object holds
raw data, normalised data, coordinates, and metadata. All analysis
functions accept the object as the first argument and return a modified
copy, enabling clean pipelines.

## Workflow at a Glance

```
ReadAkoya() -> QCFilter() -> NormaliseData() -> PhenotypeCells()
  -> FindNeighbours() / CellDensity() / InteractionMatrix()
  -> CellMap() / FeaturePlot() / ViolinPlot() / DotPlot()
  -> CompositionPlot() / SpatialNetworkPlot()
```

## Simulating Example Data

```{r simulate}
library(phenoscapR)

set.seed(42)
n <- 500

# Two spatially distinct populations
sim <- data.frame(
  `Cell ID` = seq_len(n),
  `Cell X Position` = c(rnorm(250, 200, 50), rnorm(250, 600, 50)),
  `Cell Y Position` = c(rnorm(250, 300, 50), rnorm(250, 300, 50)),
  `Cell Area (px)` = rlnorm(n, log(100), 0.3),
  CD3  = c(rnorm(250, 800, 150), rnorm(250, 200, 100)),
  CD8  = c(rnorm(250, 200, 100), rnorm(250, 700, 120)),
  PanCK = c(rnorm(250, 100, 50), rnorm(250, 600, 100)),
  check.names = FALSE
)

tmp <- tempfile(fileext = ".csv")
write.csv(sim, tmp, row.names = FALSE)
```

## Reading Data

`ReadAkoya()` reads Akoya cell segmentation CSV files and returns an
`AkoyaExperiment` object. It supports three formats: `"processor"`,
`"inform"`, and `"qupath"`, plus automatic detection (`"auto"`).

A `filter` argument removes housekeeping channels (DAPI, Blank, Empty) by default.

```{r read}
obj <- ReadAkoya(tmp, sample_id = "example", filter = NA)
obj
```

To read a directory of QuPath exports:

```{r read-dir, eval = FALSE}
# Single directory (recursive)
obj <- ReadAkoya("data/Annotations/", type = "qupath")

# Or pass a specific file
obj <- ReadAkoya("data/measurements.csv", type = "qupath")
```

## Colour Palettes

Set a global palette that all plots will use. Supports viridis palettes
and custom gradients from 2-3 anchor colours.

```{r palette}
# Use viridis (default)
SetPalette("viridis")

# Or a custom gradient
SetPalette(c("navy", "white", "firebrick"))

# Generate discrete colours for 5 groups
PaletteDiscrete(5)

# Create a reusable gradient function
pal <- CustomGradient(c("#1b9e77", "#ffffff", "#d95f02"))
pal(10)

# Reset to viridis for the rest of the vignette
SetPalette("viridis")
```

## Quality Control

```{r qc}
obj <- QCFilter(obj, min_area = 50, max_area = 500)
NCells(obj)
```

## Normalisation

`NormaliseData()` normalises marker intensities and stores the result in
the `data` slot. The raw `counts` remain untouched.

```{r normalise}
obj <- NormaliseData(obj, method = "zscore")
head(GetData(obj)[, 1:3])
```

## Phenotyping

```{r phenotype}
obj <- PhenotypeCells(obj, thresholds = list(CD3 = 0, CD8 = 0, PanCK = 0))
PhenotypeSummary(obj)
```

## Spatial Analysis

### Nearest Neighbours

```{r nn}
obj <- FindNeighbours(obj, k = 5)
summary(Meta(obj)$nn_distance)
```

### Cell Density

```{r density}
obj <- CellDensity(obj, radius = 50)
summary(Meta(obj)$density)
```

### Interaction Matrix

```{r interactions}
interactions <- InteractionMatrix(obj, radius = 50)
interactions
```

### Neighbourhood Enrichment

Permutation-based test for spatial enrichment/depletion of phenotype pairs:

```{r enrichment}
ne <- NeighbourhoodEnrichment(obj, radius = 50, n_perm = 50, seed = 42)
ne[order(ne$z_score), ]
```

### Ripley's K and Pair Correlation

```{r ripleys}
kf <- RipleysK(obj)
head(kf)
```

### Moran's I

```{r morans}
mi <- MoransI(obj, feature = "CD3", radius = 80)
cat("Moran's I:", round(mi$I, 3), "  p-value:", signif(mi$p_value, 3), "\n")
```

### Quadrat Analysis

```{r quadrat}
qa <- QuadratAnalysis(obj, nx = 5, ny = 5)
cat("VMR:", round(qa$VMR, 2), "  p-value:", signif(qa$p_value, 3), "\n")
```

### Delaunay Network

```{r delaunay}
obj <- DelaunayNetwork(obj, max_edge = 100)
head(obj@spatial$delaunay_edges)
```

### Expression Clustering

```{r expr-cluster}
obj <- ExpressionClusters(obj, k = 3)
table(Meta(obj)$expr_cluster)
```

## Visualisation

### Cell Map

```{r cell-map, fig.cap="Cell phenotype map"}
CellMap(obj, pt_size = 1)
```

### Cell Map (dark theme)

```{r cell-map-dark, fig.cap="Dark tissue-style phenotype map"}
CellMap(obj, pt_size = 1, dark_theme = TRUE)
```

### Feature Plot

```{r feature-plot, fig.cap="Marker expression in tissue space"}
FeaturePlot(obj, features = c("CD3", "CD8", "PanCK"), ncol = 3)
```

### Density Plot

```{r density-plot, fig.cap="Cell density map"}
DensityPlot(obj, pt_size = 1)
```

### Spatial Network Plot

```{r network-plot, fig.cap="Delaunay network coloured by phenotype"}
SpatialNetworkPlot(obj, pt_size = 1.5)
```

### Violin Plot

```{r violin, fig.cap="Marker distributions by phenotype"}
ViolinPlot(obj, features = c("CD3", "CD8", "PanCK"))
```

### Box Plot

```{r boxplot, fig.cap="Marker box plots"}
BoxPlot(obj, features = c("CD3", "CD8"))
```

### Dot Plot

```{r dotplot, fig.cap="Dot plot of marker expression"}
DotPlot(obj, features = c("CD3", "CD8", "PanCK"))
```

### Composition Plot

```{r composition, fig.cap="Phenotype composition"}
CompositionPlot(obj)
```

### Ridge Plot

```{r ridge, fig.cap="Ridge plot of CD3 expression"}
RidgePlot(obj, features = "CD3")
```

### Marker Heatmap

```{r heatmap, fig.cap="Mean marker expression per phenotype"}
MarkerHeatmap(obj)
```

### Interaction Heatmap

```{r interaction-plot, fig.cap="Spatial interaction scores"}
InteractionPlot(interactions)
```

### Histogram

```{r histogram, fig.cap="CD3 expression histogram"}
HistogramPlot(obj, feature = "CD3", group_by = "phenotype")
```

## Pipe-friendly Workflow

With R 4.1+ you can use the native pipe:

```{r pipe, eval = FALSE}
obj <- ReadAkoya("segmentation.csv", type = "qupath") |>
  QCFilter(min_area = 50) |>
  NormaliseData(method = "zscore") |>
  PhenotypeCells(thresholds = list(CD3 = 0.5, CD8 = 0.3)) |>
  FindNeighbours(k = 5) |>
  CellDensity(radius = 50) |>
  DelaunayNetwork(max_edge = 100)
```

## Subsetting

Subset cells or markers using standard R bracket syntax:

```{r subset}
# First 100 cells
sub <- obj[1:100, ]
NCells(sub)

# Select specific markers
sub2 <- obj[, c("CD3", "CD8")]
Markers(sub2)
```

```{r cleanup, include = FALSE}
unlink(tmp)
```
