---
title: "Getting Started with phenoscapR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with phenoscapR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 5
)
```

## Overview

**phenoscapR** provides an S4 class (`AkoyaExperiment`) and a pipe-friendly
workflow for analysing multiplexed spatial biology data from Akoya
Biosciences platforms (PhenoCycler, CODEX, PhenoImager).

The design follows the philosophy of the
[Seurat](https://satijalab.org/seurat/) package: a central object holds
raw data, normalised data, coordinates, and metadata. All analysis
functions accept the object as the first argument and return a modified
copy, enabling clean pipelines.

## Workflow at a Glance

```
ReadAkoya() → QCFilter() → NormaliseData() → PhenotypeCells()
  → FindNeighbours() / CellDensity() / InteractionMatrix()
  → CellMap() / DensityPlot() / MarkerHeatmap() / InteractionPlot()
```

## Simulating Example Data

```{r simulate}
library(phenoscapR)

set.seed(42)
n <- 500

# Two spatially distinct populations
sim <- data.frame(
  `Cell ID` = seq_len(n),
  `Cell X Position` = c(rnorm(250, 200, 50), rnorm(250, 600, 50)),
  `Cell Y Position` = c(rnorm(250, 300, 50), rnorm(250, 300, 50)),
  `Cell Area (px)` = rlnorm(n, log(100), 0.3),
  CD3  = c(rnorm(250, 800, 150), rnorm(250, 200, 100)),
  CD8  = c(rnorm(250, 200, 100), rnorm(250, 700, 120)),
  DAPI = rnorm(n, 1000, 200),
  check.names = FALSE
)

tmp <- tempfile(fileext = ".csv")
write.csv(sim, tmp, row.names = FALSE)
```

## Reading Data

`ReadAkoya()` reads Akoya cell segmentation CSV files and returns an
`AkoyaExperiment` object. Column names are auto-detected and
standardised.

```{r read}
obj <- ReadAkoya(tmp, sample_id = "example")
obj
```

## Quality Control

```{r qc}
obj <- QCFilter(obj, min_area = 50, max_area = 500)
NCells(obj)
```

## Normalisation

`NormaliseData()` normalises marker intensities and stores the result in
the `data` slot. The raw `counts` remain untouched.

```{r normalise}
obj <- NormaliseData(obj, method = "zscore")
head(GetData(obj)[, 1:3])
```

## Phenotyping

```{r phenotype}
obj <- PhenotypeCells(obj, thresholds = list(CD3 = 0, CD8 = 0))
PhenotypeSummary(obj)
```

## Spatial Analysis

### Nearest Neighbours

```{r nn}
obj <- FindNeighbours(obj, k = 5)
summary(Meta(obj)$nn_distance)
```

### Cell Density

```{r density}
obj <- CellDensity(obj, radius = 50)
summary(Meta(obj)$density)
```

### Interaction Matrix

```{r interactions}
interactions <- InteractionMatrix(obj, radius = 50)
interactions
```

## Visualisation

### Cell Map

```{r cell-map, fig.cap="Cell phenotype map"}
CellMap(obj, pt_size = 1)
```

### Density Plot

```{r density-plot, fig.cap="Cell density map"}
DensityPlot(obj, pt_size = 1)
```

### Marker Heatmap

```{r heatmap, fig.cap="Mean marker expression per phenotype"}
MarkerHeatmap(obj)
```

### Interaction Heatmap

```{r interaction-plot, fig.cap="Spatial interaction scores"}
InteractionPlot(interactions)
```

## Pipe-friendly Workflow

With R 4.1+ you can use the native pipe:

```{r pipe, eval = FALSE}
obj <- ReadAkoya("segmentation.csv") |>
  QCFilter(min_area = 50) |>
  NormaliseData(method = "zscore") |>
  PhenotypeCells(thresholds = list(CD3 = 0.5, CD8 = 0.3)) |>
  FindNeighbours(k = 5) |>
  CellDensity(radius = 50)
```

## Subsetting

Subset cells or markers using standard R bracket syntax:

```{r subset}
# First 100 cells
sub <- obj[1:100, ]
NCells(sub)

# Select specific markers
sub2 <- obj[, c("CD3", "CD8")]
Markers(sub2)
```

```{r cleanup, include = FALSE}
unlink(tmp)
```
