---
title: "phenoscapR -- Function Showcase"
subtitle: "Complete walkthrough of all phenoscapR functions using real tFL data"
author: "phenoscapR contributors"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    theme: cosmo
    self-contained: true
execute:
  warning: false
  message: false
---

## Setup

```{r}
#| label: setup
library(phenoscapR)
library(data.table)
library(ggplot2)
```

## 1. Reading Data -- `read_akoya()`

The package reads Akoya cell segmentation CSV files and auto-detects column naming
conventions (PhenoCycler, QuPath, CODEX formats).

### 1.1 Read a single sample

```{r}
#| label: read-single
tfl1 <- read_akoya(
  path = "../_data/Annotations transformierte FL/Annotations tFL1/tFL1.csv",
  sample_id = "tFL1"
)

cat("Dimensions:", nrow(tfl1), "cells x", ncol(tfl1), "columns\n")
cat("Columns:", paste(names(tfl1), collapse = ", "), "\n")
head(tfl1)
```

### 1.2 Read multiple samples from a directory

We read two samples (tFL1 and tFL2) by placing their CSVs in a temporary
directory -- demonstrating the batch-read capability.

```{r}
#| label: read-multi
# Read tFL2 separately to combine later
tfl2 <- read_akoya(
  path = "../_data/Annotations transformierte FL/Annotations tFL2/tFL2.csv",
  sample_id = "tFL2"
)

# Combine into a multi-sample dataset
dt_multi <- rbindlist(list(tfl1, tfl2))
cat("Combined:", nrow(dt_multi), "cells across",
    length(unique(dt_multi$sample_id)), "samples\n")
table(dt_multi$sample_id)
```

### 1.3 Read with marker selection

Select only specific markers of interest to reduce data width.

```{r}
#| label: read-markers
tfl1_subset <- read_akoya(
  path = "../_data/Annotations transformierte FL/Annotations tFL1/tFL1.csv",
  sample_id = "tFL1",
  markers = c("CD3e: Cell: Mean", "CD20: Cell: Mean", "Ki67: Cell: Mean")
)
cat("Selected columns:", paste(names(tfl1_subset), collapse = ", "), "\n")
```


## 2. Quality Control -- `qc_filter()`

Filter cells based on area thresholds and total marker intensity.
Since the tFL data has no `cell_area` column, we demonstrate intensity-based filtering.

```{r}
#| label: qc-filter
cat("Before QC:", nrow(tfl1), "cells\n")

# Remove cells with very low total marker intensity (likely debris)
# and extremely high intensity (likely artefacts)
tfl1_qc <- qc_filter(
  tfl1,
  min_intensity = 100,
  max_intensity = 50000
)

cat("After QC:", nrow(tfl1_qc), "cells\n")
cat("Removed:", nrow(tfl1) - nrow(tfl1_qc), "cells\n")
```


## 3. Normalisation -- `normalise_markers()`

Three normalisation methods are available: z-score, min-max, and quantile.

### 3.1 Z-score normalisation (default)

```{r}
#| label: norm-zscore
tfl1_norm <- normalise_markers(tfl1_qc, method = "zscore")

# Verify: z-scored columns should have mean ~ 0
marker_cols <- setdiff(
  names(tfl1_norm)[sapply(tfl1_norm, is.numeric)],
  c("cell_id", "x", "y")
)
cat("Mean of first 3 markers after z-score:\n")
sapply(marker_cols[1:3], function(m) round(mean(tfl1_norm[[m]], na.rm = TRUE), 4))
```

### 3.2 Min-max normalisation

```{r}
#| label: norm-minmax
tfl1_mm <- normalise_markers(tfl1_qc, method = "minmax")

# Values should be in [0, 1]
cat("Range of CD3e after minmax:",
    range(tfl1_mm[["CD3e: Cell: Mean"]], na.rm = TRUE), "\n")
```

### 3.3 Quantile normalisation

```{r}
#| label: norm-quantile
tfl1_qt <- normalise_markers(tfl1_qc, method = "quantile")

# Values should be in [0, 1] with uniform distribution
cat("Summary of CD20 after quantile normalisation:\n")
summary(tfl1_qt[["CD20: Cell: Mean"]])
```


## 4. Phenotyping -- `phenotype_cells()`

Assign phenotype labels based on marker intensity thresholds.
We use the z-score normalised data and define biologically meaningful thresholds.

```{r}
#| label: phenotype
tfl1_pheno <- phenotype_cells(
  tfl1_norm,
  thresholds = list(
    "CD3e: Cell: Mean" = 1.0,   # T cells
    "CD20: Cell: Mean" = 1.0,   # B cells
    "CD68: Cell: Mean" = 1.0    # Macrophages
  )
)

cat("Phenotype distribution:\n")
table(tfl1_pheno$phenotype)
```

### Custom labels

```{r}
#| label: phenotype-labels
tfl1_labelled <- phenotype_cells(
  tfl1_norm,
  thresholds = list(
    "CD3e: Cell: Mean" = 1.0,
    "CD20: Cell: Mean" = 1.0
  ),
  labels = c(
    "CD3e: Cell: Mean+" = "T cells",
    "CD20: Cell: Mean+" = "B cells",
    "CD3e: Cell: Mean+/CD20: Cell: Mean+" = "Double positive",
    "Negative" = "Other"
  )
)
table(tfl1_labelled$phenotype)
```


## 5. Phenotype Summary -- `summarise_phenotypes()`

Compute phenotype frequencies and proportions per sample.

```{r}
#| label: summarise-pheno
# First phenotype both samples
tfl2_qc <- qc_filter(tfl2, min_intensity = 100, max_intensity = 50000)
tfl2_norm <- normalise_markers(tfl2_qc, method = "zscore")
tfl2_pheno <- phenotype_cells(
  tfl2_norm,
  thresholds = list(
    "CD3e: Cell: Mean" = 1.0,
    "CD20: Cell: Mean" = 1.0,
    "CD68: Cell: Mean" = 1.0
  )
)

dt_combined <- rbindlist(list(tfl1_pheno, tfl2_pheno))
pheno_summary <- summarise_phenotypes(dt_combined)
print(pheno_summary)
```

```{r}
#| label: summarise-pheno-plot
#| fig-width: 8
#| fig-height: 5
ggplot(pheno_summary, aes(x = sample_id, y = proportion, fill = phenotype)) +
  geom_col(position = "stack") +
  theme_minimal() +
  labs(x = "Sample", y = "Proportion", fill = "Phenotype",
       title = "Phenotype proportions per sample")
```


## 6. Spatial Analysis

### 6.1 Nearest Neighbours -- `nearest_neighbours()`

Compute the mean distance to the k nearest neighbours for each cell.

```{r}
#| label: nn
# Use a subsample for speed (tFL1 has ~14k cells)
set.seed(42)
tfl1_sub <- tfl1_pheno[sample(.N, min(.N, 2000))]

tfl1_nn <- nearest_neighbours(tfl1_sub, k = 5)
cat("nn_distance summary:\n")
summary(tfl1_nn$nn_distance)
```

#### Nearest neighbour distance to a specific phenotype

```{r}
#| label: nn-target
tfl1_nn_cd20 <- nearest_neighbours(
  tfl1_sub,
  target_phenotype = "CD20: Cell: Mean+",
  k = 3
)
cat("Distance to nearest B cells:\n")
summary(tfl1_nn_cd20$nn_distance)
```


### 6.2 Cell Density -- `cell_density()`

Count the number of neighbours within a given radius.

```{r}
#| label: density
tfl1_dens <- cell_density(tfl1_sub, radius = 50)
cat("Local density summary (neighbours within 50 units):\n")
summary(tfl1_dens$density)
```

#### Density of a specific phenotype

```{r}
#| label: density-target
tfl1_dens_t <- cell_density(
  tfl1_sub,
  radius = 50,
  target_phenotype = "CD3e: Cell: Mean+"
)
cat("T cell density around each cell:\n")
summary(tfl1_dens_t$density)
```


### 6.3 Interaction Matrix -- `interaction_matrix()`

Compute pairwise spatial interaction scores between phenotypes.
Positive scores indicate spatial co-localisation (attraction),
negative scores indicate avoidance.

```{r}
#| label: interactions
interactions <- interaction_matrix(tfl1_sub, radius = 50)
print(interactions)
```


### 6.4 Spatial Clustering -- `spatial_clusters()`

Cluster cells based on their spatial (x, y) coordinates.

#### K-means clustering

```{r}
#| label: cluster-kmeans
tfl1_clust <- spatial_clusters(tfl1_sub, k = 4, method = "kmeans")
table(tfl1_clust$cluster)
```

#### Hierarchical clustering

```{r}
#| label: cluster-hclust
tfl1_hclust <- spatial_clusters(tfl1_sub, k = 4, method = "hierarchical")
table(tfl1_hclust$cluster)
```


## 7. Visualisation

### 7.1 Cell Map -- `plot_cell_map()`

Scatter plot of cell positions coloured by phenotype.

```{r}
#| label: fig-cellmap-phenotype
#| fig-width: 9
#| fig-height: 7
plot_cell_map(tfl1_pheno, colour_by = "phenotype",
              point_size = 0.3,
              title = "tFL1 -- Cell phenotype map")
```

#### Colour by spatial cluster

```{r}
#| label: fig-cellmap-cluster
#| fig-width: 9
#| fig-height: 7
plot_cell_map(tfl1_clust, colour_by = "cluster",
              point_size = 0.8,
              title = "tFL1 subsample -- Spatial clusters (k-means, k=4)")
```

#### Custom colours

```{r}
#| label: fig-cellmap-custom
#| fig-width: 9
#| fig-height: 7
plot_cell_map(
  tfl1_labelled,
  colour_by = "phenotype",
  colours = c(
    "T cells" = "#E41A1C",
    "B cells" = "#377EB8",
    "Double positive" = "#984EA3",
    "Other" = "grey80"
  ),
  point_size = 0.3,
  title = "tFL1 -- Custom phenotype colours"
)
```


### 7.2 Density Map -- `plot_density()`

Visualise local cell density across the tissue.

```{r}
#| label: fig-density
#| fig-width: 9
#| fig-height: 7
plot_density(tfl1_dens, point_size = 0.8,
             title = "tFL1 subsample -- Local cell density (radius=50)")
```


### 7.3 Marker Heatmap -- `plot_heatmap()`

Mean marker intensity per phenotype displayed as a heatmap.

```{r}
#| label: fig-heatmap
#| fig-width: 10
#| fig-height: 5
plot_heatmap(tfl1_pheno,
             title = "tFL1 -- Mean marker intensity per phenotype")
```

#### Heatmap with selected markers

```{r}
#| label: fig-heatmap-select
#| fig-width: 8
#| fig-height: 5
plot_heatmap(
  tfl1_pheno,
  markers = c("CD3e: Cell: Mean", "CD20: Cell: Mean",
              "CD68: Cell: Mean", "Ki67: Cell: Mean",
              "Bcl-2: Cell: Mean"),
  title = "tFL1 -- Selected markers"
)
```


### 7.4 Interaction Heatmap -- `plot_interactions()`

Visualise spatial interaction scores between phenotypes.

```{r}
#| label: fig-interactions
#| fig-width: 7
#| fig-height: 6
plot_interactions(interactions,
                  title = "tFL1 subsample -- Spatial interaction scores")
```


## 8. Full Pipeline Example

Putting it all together in a single streamlined workflow.

```{r}
#| label: full-pipeline
# 1. Read
dat <- read_akoya(
  "../_data/Annotations transformierte FL/Annotations tFL1/tFL1.csv",
  sample_id = "tFL1"
)

# 2. QC
dat <- qc_filter(dat, min_intensity = 100, max_intensity = 50000)

# 3. Normalise
dat <- normalise_markers(dat, method = "zscore")

# 4. Phenotype
dat <- phenotype_cells(dat, thresholds = list(
  "CD3e: Cell: Mean" = 1.0,
  "CD20: Cell: Mean" = 1.0,
  "CD68: Cell: Mean" = 1.0
))

# 5. Summary
summarise_phenotypes(dat)

# 6. Spatial analysis (subsample for speed)
set.seed(123)
dat_sub <- dat[sample(.N, min(.N, 2000))]
dat_sub <- nearest_neighbours(dat_sub, k = 5)
dat_sub <- cell_density(dat_sub, radius = 50)
int <- interaction_matrix(dat_sub, radius = 50)
dat_sub <- spatial_clusters(dat_sub, k = 3)
```

```{r}
#| label: fig-pipeline-panel
#| fig-width: 12
#| fig-height: 10
#| layout-ncol: 2
# 7. Visualise
plot_cell_map(dat, colour_by = "phenotype", point_size = 0.2,
              title = "Phenotype map")
plot_density(dat_sub, point_size = 0.8,
             title = "Cell density")
plot_heatmap(dat, title = "Marker heatmap")
plot_interactions(int, title = "Spatial interactions")
```


## Session Info

```{r}
#| label: session-info
sessionInfo()
```
